Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-09-20 Tex/Paper.tex
Implemented many of Patric's comments

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativistic beam travelling through a
plasma excites a strong longitudinal e-field that can be loaded by a trailing witness beam. With optimal beam loading,
an increase in energy spread can be kept to a minimum. Acceleration of an electron withness beam by an electron drive
beam has been demonstrated experimentally in the past \cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at
CERN is a proof of concept proton driven plasma wakefield accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, the non-linear transverse wakefields cause emittance growth of the accelerated
beam. The beam will also see a transversely varying accelerating field causing increasing energy spread
\cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by the transverse
oscillations of the plasma electrons forming a sheet around an evacuated area leaving only ions. The ions, assumed
stationary, form an ion channel creating a focusing force that varies linearly with radius. This produces an axially
symmetric focusing force \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
%~ \begin{align}
    %~ \lambda_{p} = 2\pi\frac{c}{\omega_{pe}} \label{EQ:LambdaP}
%~ \end{align}
%~ and
%~ \begin{align}
    %~ \omega_{pe} = \sqrt{\frac{n_{0} e^{2}}{m_{e} \epsilon_{0}}}, \label{EQ:OmegaP}
%~ \end{align}
$\lambda_{p} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$,
will produce a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness beam loading the
peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches. Acceleration of
an electron witness beam driven by two electron drive beams has been demonstrated at Brookhaven National Laboratory
\cite{muggli:2011}.

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. The energy of a
high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam in the international linear collider with
$1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers as drivers, a large number of plasma stages
is required. How to efficiently stage plasma accelerators is challenging without reducing the effective gradient and
spoiling the beam quality \cite{steinke:2016,lindstrom:2016}. Proton drive beams available at CERN carry significantly
more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for longer accelerator stages. However,
the SPS beam is orders of magnitude longer than the plasma wavelengths needed for such applications. This issue is
resolved by letting the proton beam undergo self-modulation before injecting the witness beam into one of the buckets in
the modulated structure. This self-modulation instability is caused by the transverse fields generated by the beam
acting upon the beam itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation frequency
is close to that of the plasma, and leaves a train of proton bunches along the beam axis with a surrounding halo.
%~ De-phasing of the
%~ witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
%~ beam, limits the effective length of a plasma accelerator stage.

@@

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver, and a nominal plasma density $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam
such that $k_{pe}\sigma_{x} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first
phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the
wakefield with a long electron beam. The study presented here is for Run 2 \cite{adli:2016}, which aims to demonstrate
acceleration of a short electron beam to high energy with a minimal increase in emittance and energy spread.

@@

The preliminary design of Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without
the electron beam present. The electron witness beam will then be injected into the modulated proton beam before stage
two, where it undergoes acceleration.
%~ As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
%~ keep this as short as possible.

With this experimental design the self-modulated proton beam does not produce a full non-linear wakefield, and not all
plasma electrons are therefore evacuated. The result is that the focusing force will not increase linearly with radius.
The small size of the accelerating structure produced by the self-modulated beam also puts constraints on the size of
the electron witness beam, and thus its charge and current -- if we at the same time wish to prevent large energy
spread. By matching the transverse size of the beam at a given emittance to the plasma density, we prevent large
amplitude oscillations which may cause additional growth in energy spread as well as in emittance.

The key idea in this paper is that while the proton beam wake is not fully blown-out leading to non-linear focusing
forces, the electron beam, if intense enough, may load the field and at the same time create its own ion bubble. As will
be shown, exploiting these two well known effect, the electron beam can be accelerated, for long distances, without
significant emittance growth.

@@

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors
that may affect this (add examples?), we have tried several approaches to create a stable drive beam structure based on
previous self-modulation studies.

@@

In order to study the witness beam emittance evolution we turned to the recently released open source version of
QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D quasi-static PIC code \cite{huang:2006, an:2013}.
Since QuickPIC is a quasi-static code, it does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, making it a well suited code to study emittance preservation.

@@

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
also set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. In order to create a stable environment to study
the evolution of the witness beam we in addition prevented the proton beam from evolving radially by increasing the
particle mass by a factor of $1\mexp{6}$. We are not considering the evolution of the proton beam itself in this study.

@@

The witness beam in our simulations differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. The relative drift causing the de-phasing is proportional to $\gamma^{-2}$. By simply matching the
initial $\gamma$ values we eliminate this effect entirely, but a lower initial value is likely to be sufficient
\cite{berglyd_olsen:2015}.

In these 3D simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{x}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_{pe}}{2\pi}\sqrt{2\gamma}, \label{EQ:Matched}
\end{align}
where $\lambda_{pe}$ is the plasma wavelength and $\gamma$ is the beam relativistic factor. This relation requires a
very narrow beam compared to the drive beam $\sigma_{x} = 200\unit{\mu m}$. The charge density of this compact witness
beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal beam loading at around
$1-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear wakefield already at the
head of the beam. The majority of the electrons within it will therefore see a linear focusing force preventing the bulk
of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam. We used a transverse grid cell size of $1.17\unit{\mu m}$, and $2.34\unit{\mu m}$ for
the longitudinal grid cells. The witness beam was simulated with $16.8\unit{M}$ non-weighted particles. For the largest
parameter scans, shown in figure \ref{Fig:BeamQ}, we used a transverse grid cell size of $2.34\unit{\mu m}$.

In the following section we study a witness beam with a total charge of $100\unit{pC}$, a $\sigma_{z}=60\unit{\mu m}$,
and with a $\sigma_{x}=5.23\unit{\mu m}$ matching an initial normalised emittance $\epsilon_{0} = 2\unit{\mu m}$. We
refer to this parameter set as our base case.

@@

The single drive beam setup is designed to behave similarly to the self-modulated case. However, since the drive beam is
prevented from significant transverse evolution, we are presented with an idealised case where the electron witness beam
sees consistent wakefields throughout the plasma stage. The $E_{z}$-field generated by the proton drive beam is seen as
the blue line in figure \ref{Fig:BeamLoading}, shown with and without the electron beam present. With a proton beam
density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime but near non-linear \cite{rosenzweig:2010}. The dashed
green line in the lower part of figure \ref{Fig:BeamLoading} shows that the on-axis plasma density has a depletion of
$67\%$, close to what we see in full scale reference simulations for AWAKE Run 2 \cite{awake_collaboration:2016}.

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    %~ \includegraphics[height=22.2mm,trim={ 0mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation01}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation02}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  0mm 0mm},clip]{figures/beamFilamentation03}
    %~ \includegraphics[height=22.2mm,trim={10mm 9.5mm 15mm 0mm},clip]{figures/beamOffsetMean}
    %~ \\
    %~ \includegraphics[height=26.0mm,trim={ 0mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation04}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation05}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  0mm 2mm},clip]{figures/beamFilamentation06}
    %~ \includegraphics[height=26.0mm,trim={10mm 0.1mm 15mm 2mm},clip]{figures/beamOffsetRMS}
    \caption{\label{Fig:BeamFilament} The left plots show the transverse phase space of the electron beam at different
        plasma positions. The Right side shows the macro particle mean position (top) and RMS spread (bottom). The blue
        particles and lines represent particles with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. The red
        particles and lines represent particles with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

@@

The charge density required to optimally load the $E_{z}$ field is $n_{eb} \approx 35\times n_{0}$. This means that the
witness beam's own wakefield is in the fully non-linear regime, where the space charge force is sufficient to blow-out
all plasma electrons resulting in an ion column. This ion column, as is well known \cite{rosenzweig:1991}, provides a
linear focusing force on the part the electron beam within the column, and therefore prevents emittance growth for this
region of the beam. This effect is shown for our base case in figure \ref{Fig:PlasmaDenTWake}. The focusing field has a
gradient of $20\unit{kT/m}$ near the beam axis.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied that maintained a stable accelerating structure, about $70-80\%$ of the beam
retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows emittance growth for the base case as a function of
propagation length. No significant emittance growth was observed for propagation lengths up to $100\unit{m}$. The drive
beam energy was here increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing, as de-phasing starts to become a
significant effect for the SPS beam after about $50\unit{m}$.

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius follows the evolution given by equation \ref{EQ:Matched} with $\lambda_{pe} = \lambda_{0}$ for the section
of the beam where normalised emittance is preserved -- as shown in figure \ref{Fig:BeamFilament}. This has the potential
to cause overloading of the field. However, for our base case this effect was not significant.

@@

The authors would also like to acknowledge the OSIRIS Consortium for providing access to the OSIRIS framework. OSIRIS
was used extensively for simulations leading up to the work presented in this paper.

>>>>

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativistic charged particle beam
travelling through a plasma excites a strong longitudinal electric field that can be loaded by a trailing witness beam.
With optimal beam loading, an increase in energy spread can be kept to a minimum. Acceleration of an electron withness
beam by an electron drive beam has been demonstrated experimentally in the past
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at CERN is a proof of concept proton driven plasma wakefield
accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the
accelerated beam. The beam will also see a transversely and longitudinally varying accelerating field causing an
increase in energy spread \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons forming a sheet around an evacuated area filled with only ions. The
ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly with
radius. This produces an axially symmetric focusing force which preserves emittance \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
%~ \begin{align}
    %~ \lambda_{pe} = 2\pi\frac{c}{\omega_{pe}} \label{EQ:LambdaP}
%~ \end{align}
%~ and
%~ \begin{align}
    %~ \omega_{pe} = \sqrt{\frac{n_{0} e^{2}}{m_{e} \epsilon_{0}}}, \label{EQ:OmegaP}
%~ \end{align}
$\lambda_{pe} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$,
produces a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness beam loading the
peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches when they are
properly phased with respect to the wakefields. Acceleration of an electron witness beam driven by two electron drive
beams has been demonstrated at Brookhaven National Laboratory \cite{muggli:2011}.

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. The energy of a
high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam in the international linear collider with
$1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers pulses as drivers, a large number of plasma
stages is required. Staging plasma accelerators without reducing the effective gradient and spoiling the beam quality is
challenging \cite{steinke:2016,lindstrom:2016}. Proton beams available at CERN carry significantly more energy,
$19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for driving longer plasma wakefield accelerator
stages. However, the SPS beam is orders of magnitude longer than the plasma wavelengths needed for such applications.
This issue is resolved by letting the proton beam undergo self-modulation before injecting the witness beam into one of
the buckets in the modulated structure. This self-modulation process is caused by the transverse fields generated by the
beam acting upon the beam itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation
frequency is close to that of the plasma, and produces a train of short proton bunches along the beam axis with a
surrounding halo of defocused particles. The train resonantly produces wakefields to larger amplitudes.
%~ De-phasing of the
%~ witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
%~ beam, limits the effective length of a plasma accelerator stage.

@@

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver and a nominal plasma density of $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the transverse size
of the SPS proton beam such that $k_{pe}\sigma_{x} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number.
The aim of the first phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam. The aim of
the second phase in 2018 is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented
here is for Run 2 \cite{adli:2016}, which aims to demonstrate acceleration of a short electron beam, $\ll\lambda_{pe}$,
to high energy with a minimal increase in emittance and energy spread.

@@

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}. The first
section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the electron
beam present. The electron witness beam is then  injected into the modulated proton beam before stage two, where it
undergoes acceleration.
%~ As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
%~ keep this as short as possible.

Using the SPS proton beam as driver, the self-modulated proton beam does not produce a full non-linear wakefield, and
not all plasma electrons are therefore evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius. The small size of the accelerating structure produced by the self-modulated beam also
puts constraints on the size of the electron witness beam, and thus its charge and current -- if at the same time we
wish to prevent large energy spread. By matching the transverse size of the beam at a given emittance to the plasma
density, we prevent large amplitude oscillations which may cause additional growth in energy spread as well as in
emittance.

The key idea in this paper is that while the proton beam wake is not fully blown-out, and thus leading to non-linear
focusing forces, the electron beam, if intense enough, may load the field and at the same time create its own ion
bubble. As will be shown, exploiting these two well known effects, the electron beam can be accelerated for long
distances without significant emittance growth.

@@

The main focus of this study is on the loading of the wakefields driven by the proton beam. In order to eliminate other
factors that may affect this (add examples?), we tried several approaches to create a stable drive beam structure based
on previous self-modulation studies.

@@

In order to study the witness beam emittance evolution we turned to the recently released open source version of
QuickPIC. QuickPIC is a fully relativistic 3D quasi-static PIC code \cite{huang:2006, an:2013}. It does not suffer from
the numerical Cherenkov effect that full PIC codes do \cite{godfrey:1974,lehe:2013}, making it a well suited tool to
study emittance preservation.

@@

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
also set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. We are not considering the evolution of the
proton beam itself in this study, so in order to create a stable environment to study the evolution of the witness beam
we prevent the proton beam from evolving radially by increasing the particle mass 6 orders of magnitude.

@@

Beam loading of a short witness beam is sensitive to its position relative to the field \cite{tzoufras:2009}. To
eliminate the problem of initial de-phasing of the witness beam with respect to the proton wakefield, initial beam
energy was set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. The relative drift causing the de-phasing is proportional
to $\gamma^{-2}$. By simply matching the initial $\gamma$ values we eliminate this effect entirely. However, a lower
initial value is likely to be sufficient \cite{berglyd_olsen:2015}.

In these 3D simulations we have used a matched beam for an initial normalised emittance of $\epsilon = 2\unit{\mu m}$,
corresponding to a $\sigma_{x}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2\gamma}{\epsilon} = \frac{\lambda_{pe}}{2\pi}\sqrt{2\gamma}, \label{EQ:Matched}
\end{align}
This relation requires a very narrow beam compared to the drive beam $\sigma_{x} = 200\unit{\mu m}$. The charge density
of this compact witness beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal beam
loading at around $1-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear
wakefield already at the head of the beam. The majority of the electrons within it will therefore see a linear focusing
force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the
narrower one. We used a transverse grid cell size of $1.17\unit{\mu m}$, and of $2.34\unit{\mu m}$ for the longitudinal
grid cells. The witness beam was simulated with $16.8\unit{M}$ non-weighted particles. For the largest parameter scans,
shown in figure \ref{Fig:BeamQ}, we used a transverse grid cell size of $2.34\unit{\mu m}$.

In the following section we study a witness beam with a total charge of $100\unit{pC}$, a $\sigma_{z}=60\unit{\mu m}$,
and with a $\sigma_{x}=5.25\unit{\mu m}$ matching an initial normalised emittance $\epsilon_{0} = 2\unit{\mu m}$. We
refer to this parameter set as our base case.

@@

The single drive beam setup is designed to produce similar initial conditions for the witness beam as the self-modulated
case. However, since the drive beam is prevented from significant transverse evolution, we are presented with an
idealised case where the electron witness beam sees consistent wakefields throughout the plasma stage. The $E_{z}$-field
generated by the proton drive beam is seen as the blue line in figure \ref{Fig:BeamLoading}, shown with and without the
electron beam present. With a proton beam density $n_{pb} \simeq n_{0}$, we are in the quasi-linear regime
\cite{rosenzweig:2010}. The dashed green line in the lower part of figure \ref{Fig:BeamLoading} shows that the on-axis
plasma density has a depletion of $67\%$, close to what we see in full scale reference simulations for AWAKE Run 2
\cite{awake_collaboration:2016}.

@@

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    %~ \includegraphics[height=22.2mm,trim={ 0mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation01}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation02}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  0mm 0mm},clip]{figures/beamFilamentation03}
    %~ \includegraphics[height=22.2mm,trim={10mm 9.5mm 15mm 0mm},clip]{figures/beamOffsetMean}
    %~ \\
    %~ \includegraphics[height=26.0mm,trim={ 0mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation04}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation05}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  0mm 2mm},clip]{figures/beamFilamentation06}
    %~ \includegraphics[height=26.0mm,trim={10mm 0.1mm 15mm 2mm},clip]{figures/beamOffsetRMS}
    \caption{\label{Fig:BeamFilament} The left plots show the transverse phase space of the electron beam at different
        plasma positions. The right side shows the macro particle mean position (top) and RMS spread (bottom). The blue
        particles and lines represent particles with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. The red
        particles and lines represent particles with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

@@

The charge density required to optimally load the $E_{z}$ field is $n_{eb} \approx 35\times n_{0}$. This means that the
witness beam's own wakefield is in the fully non-linear regime, where the space charge force is sufficient to blow-out
all plasma electrons resulting in the formation of an ion column. This ion column, as is well known
\cite{rosenzweig:1991}, provides a linear focusing force on the part the electron beam within the column, and therefore
prevents emittance growth for this region of the beam. This effect is shown for our base case in figure
\ref{Fig:PlasmaDenTWake}. The focusing field has a gradient of $20\unit{kT/m}$ near the beam axis.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied that maintained a stable accelerating structure, about $70-80\%$ of the
electron beam retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows emittance growth for the base case as a
function of propagation length. No significant emittance growth was observed for propagation lengths up to
$100\unit{m}$. The drive beam energy was here increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing, as
de-phasing starts to become a significant effect for the SPS beam of $400+unit{GeV}$ after about $50\unit{m}$.

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius within the bubble, where normalised emittance is preserved, follows the evolution given by equation
\ref{EQ:Matched} -- as shown in figure \ref{Fig:BeamFilament}. This has the potential to cause overloading of the field.
However, for our base case this effect was not significant.

@@

The authors would also like to thank the OSIRIS Consortium for providing access to the OSIRIS framework. OSIRIS was used
extensively for simulations leading to the work presented in this paper.
